---
title: "Stomach Methylation - Differential Methylation Analysis v02"
author: "Hilmar Berger, based on code by Kristin Fritsche"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(ChAMP)
library(ChAMPdata)
library(matrixStats)
library(pheatmap)
library(dendextend)
library(plyr)
library(writexl)

library(limma)

result_folder = paste("./Results",format(Sys.time(), "%Y-%m-%d"),sep="/")
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

data_storage_folder = file.path("./Results/data_storage","2021-06-10")
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

```


# Experimental Design

Gastric biopsies from patients with intestinal metaplasia were taken from metaplastic and non-metaplastic control sites. Primary gastric epithelial cells were extracted and expanded on feeder cells. Single cell clones were derived, expanded on feeders and then cultured for few passages as organoid cultures to remove feeder cells. 

DNA was isolated and hybridized to the Illumina EPIC methylation array.

To analyse the generated data in R, I will use and further adapt the "Chip Analysis Methylation Pipeline" (ChAMP) package [Morris J, *et al.*, 2013] and use limma for differential methylation analysis

# Data Import

Raw intensity values were preprocessed and QC checked (see "Preprocessing_and_QC.Rmd"). Here I load the sample annotation and the normalized (SWAN) beta values:

```{r Data Import, message=FALSE}

load(file.path(data_storage_folder, "myLoad.RData"))

load(file.path(data_storage_folder, "myNorm.RData"))


library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
full_anno = getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19, "Other")

full_anno = full_anno[, c("UCSC_RefGene_Name", "UCSC_RefGene_Accession", "UCSC_RefGene_Group", "Phantom5_Enhancers", "DMR", "X450k_Enhancer", "Regulatory_Feature_Group", "GencodeCompV12_NAME", "GencodeCompV12_Accession", "GencodeCompV12_Group", "DNase_Hypersensitivity_NAME", "DNase_Hypersensitivity_Evidence_Count", "OpenChromatin_NAME", "OpenChromatin_Evidence_Count", "TFBS_NAME", "TFBS_Evidence_Count")]

```

Here, I now subset the normalized data set, in order to remove all CpGs that between all samples do not change > 20 %.

```{r}

myNorm_sub = myNorm

Diff = rowDiffs(matrixStats::rowRanges(myNorm_sub))

#myNorm_sub = cbind(myNorm_sub, Diff)
#colnames(myNorm_sub)[ncol(myNorm_sub)] = "MaxDiff"  
myNorm_sub = subset(myNorm_sub, Diff > 0.2)

## 62,574



```

# Differential Methylation Analysis

The following comparisons are going to be performed using limma [Ritchie ME, *et al*., 2015] and the subsetted data set:

- IM vs non-IM (592+594, pairwise, all replicates)
- IM vs non-IM (592+594, pairwise, replicates averaged)
- IM vs non-IM (592, pairwise, all replicates)
- IM vs non-IM (594, pairwise, all replicates)

```{r Differential Methylation Analysis}

if(FALSE) {
  
  all_results_betas = list()
  
  data_matrix = myNorm_sub
  ed = myLoad$pd
  
  
  # IM vs. IM, patients 592 and 594, all replicates as full samples
  sel_conditions = c("IM","non-IM")
  sel_samples = rownames(subset(ed, Sample_Group %in% sel_conditions & PatientID %in% c("592","594")))
  edata = data_matrix[,sel_samples]
  groups = factor(ed[sel_samples,]$Sample_Group)
  patient = factor(ed[sel_samples,]$PatientID)
  
  design = model.matrix(~ patient + groups ,  contrasts.arg=list(groups=contr.SAS))
  
  fit <- lmFit(edata,design)
  fit <- eBayes(fit)
  all_results_betas[["IMvsNonIM_592_594"]] = topTable(fit, coef="groups1", adjust="BH", number=nrow(edata))
  

  # IM vs. IM, patients 592 and 594, replicates averaged
  sel_conditions = c("IM","non-IM")
  sel_samples = rownames(subset(ed, Sample_Group %in% sel_conditions & PatientID %in% c("592","594")))
  edata = data_matrix[,sel_samples]
  groups = factor(ed[sel_samples,]$Sample_Group)
  patient = factor(ed[sel_samples,]$PatientID)

  patient_group = paste0(patient,"_",groups)
  
  edata_avg = avearrays(edata, ID=patient_group)
  groups = factor(unlist(sapply(strsplit(colnames(edata_avg),"_"), `[`,2)))
  patient = factor(unlist(sapply(strsplit(colnames(edata_avg),"_"), `[`,1)))
    
  design = model.matrix(~ patient + groups ,  contrasts.arg=list(groups=contr.SAS))
  
  fit <- lmFit(edata_avg,design)
  fit <- eBayes(fit)
  all_results_betas[["IMvsNonIM_592_594_avg"]] = topTable(fit, coef="groups1", adjust="BH", number=nrow(edata))

  
  # IM vs. IM, patient 592
  sel_conditions = c("IM","non-IM")
  sel_samples = rownames(subset(ed, Sample_Group %in% sel_conditions & PatientID %in% c("592")))
  edata = data_matrix[,sel_samples]
  groups = factor(ed[sel_samples,]$Sample_Group)
  #patient = factor(ed[sel_samples,]$PatientID)
  
  design = model.matrix(~ groups ,  contrasts.arg=list(groups=contr.SAS))
  
  fit <- lmFit(edata,design)
  fit <- eBayes(fit)
  all_results_betas[["IMvsNonIM_592"]] = topTable(fit, coef="groups1", adjust="BH", number=nrow(edata))


  # IM vs. IM, patient 594
  sel_conditions = c("IM","non-IM")
  sel_samples = rownames(subset(ed, Sample_Group %in% sel_conditions & PatientID %in% c("594")))
  edata = data_matrix[,sel_samples]
  groups = factor(ed[sel_samples,]$Sample_Group)
  #patient = factor(ed[sel_samples,]$PatientID)
  
  design = model.matrix(~ groups ,  contrasts.arg=list(groups=contr.SAS))
  
  fit <- lmFit(edata,design)
  fit <- eBayes(fit)
  all_results_betas[["IMvsNonIM_594"]] = topTable(fit, coef="groups1", adjust="BH", number=nrow(edata))


  sel_samples = colnames(data_matrix)

  groups = factor(ed[sel_samples,]$Sample_Group)
  patient = factor(ed[sel_samples,]$PatientID)

  patient_group = paste0(groups,"_",patient)

  myNorm_sub_avg = avearrays(data_matrix[,sel_samples], ID=patient_group)
  myNorm_sub_avg = myNorm_sub_avg[, order(colnames(myNorm_sub_avg))]
  colnames(myNorm_sub_avg) = paste0(colnames(myNorm_sub_avg), "_avg")
  
  #groups = factor(unlist(sapply(strsplit(colnames(),"_"), `[`,2)))
  #patient = factor(unlist(sapply(strsplit(colnames(edata_avg),"_"), `[`,1)))

  all_results_betas_annotated = list()
  for (n in names(all_results_betas)) {
    tmp = all_results_betas[[n]]
    tmp = merge(tmp, myNorm_sub_avg, by=0, all.x=T, sort=F)
    tmp = merge(tmp, as.data.frame(full_anno), by.x="Row.names", by.y=0, sort=F, all.x=T )
    all_results_betas_annotated[[n]] = tmp
  }
        
  save(all_results_betas_annotated, file=file.path(data_storage_folder, "all_results_betas_annotated.RData"))
} else {
  load(file.path(data_storage_folder, "all_results_betas_annotated.RData"))
}



```

## Volcano Plots

### Subgroups

For each comparison, the distribution of the differential methylation values (delta beta) and the negative logarithm of the adjusted p-values are plotted in the following volcano plots. The red line marks the cutoff of 0.05 of the FDR.

#### FDR < 5%

```{r Volcano Plots adj. p-val, fig.width=12,fig.height = 7.5}
all_target_conditions = names(all_results_betas_annotated)
par(mfrow=c(2,2))
for (tc in all_target_conditions) {
  r = all_results_betas_annotated[[tc]]
  plot(r$logFC, -log10(r$adj.P.Val),xlab="delta beta",ylab="-log10(adj. p-val)", ylim=c(0,max(2,max(-log10(r$adj.P.Val),na.rm=T))), xlim=(c(-1, 1)))
  title(main=tc, sub=paste("(",nrow(subset(r, adj.P.Val < 0.05 & (logFC > 0.2 | logFC < -0.2)))," diff. meth. Cgs after adjusting (FDR < 5 %, db > 20 %))",sep="") )
  abline(h=-log10(0.05),col="red")
    abline(v=-0.2, lty = 2)
  abline(v=0.2, lty = 2)
}

```

# Result files

Results have been written to file *Differential_Methylation_Results.xlsx* in `r result_folder`.

```{r}
write_xlsx(all_results_betas_annotated, path = file.path(result_folder, "Differential_Methylation_Results.xlsx"))
```


# Conclusion

Results look quite good.

Another approach might be to perform the statistical analysis with M-values. 


# Software versions

```{r}
sessionInfo()
```